---
title: "Cholera Outbreak Analysis Report"
author: "MSc Big Data Analytics Student"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)
```

# Cholera Outbreak Analysis - 2008

This report analyzes a cholera outbreak affecting two provinces. The analysis includes data cleaning, descriptive statistics, correlation analysis, and regression modeling.

---

# Load Libraries

```{r}
# Load required packages (no install calls in knit)
library(readxl)
library(ggplot2)      # visualization
library(dplyr)        # data manipulation
library(tidyr)        # tidying
library(readr)        # parse_number
library(knitr)        # reporting
library(kableExtra)   # nice tables
library(gridExtra)    # combining plots
library(reshape2)     # reshaping
library(naniar)       # vis_miss
library(viridis)      # palettes
library(lubridate)    # robust date parsing

cat("Libraries loaded successfully!\n")
```

---

# Load and Explore Data

```{r}
# Read the data (ensure the file is in the working directory)
# You can change the file name/path as needed.
chol_outbreak_df <- read_excel("DataFile.xls")

# Display first few rows
head(chol_outbreak_df, 5)
View(chol_outbreak_df)
```

---

# TASK 1: Data Quality Assessment

```{r}
# Missing values per column (NA-safe)
missing_df <- data.frame(
  Column = names(chol_outbreak_df),
  Missing_Count = sapply(chol_outbreak_df, function(v) sum(is.na(v))),
  stringsAsFactors = FALSE
) |>
  dplyr::mutate(Percent_Missing = round(100 * Missing_Count / nrow(chol_outbreak_df), 2)) |>
  dplyr::arrange(dplyr::desc(Percent_Missing))

kable(missing_df, caption = "Missing Values by Column") |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r}
# Visualize missing data pattern across all rows
naniar::vis_miss(chol_outbreak_df)
```

## Data Quality Issues Found

```{r}
# SEX VARIABLE
cat("\n\nSEX VARIABLE ISSUES:\n")
cat("===================\n")
sex_table <- table(chol_outbreak_df$Sex, useNA = "always")
print(sex_table)
cat("\nProblems: Mixed case (F, f, M, m), extra spaces, missing values\n")

# AGE VARIABLE
cat("\nAGE VARIABLE ISSUES:\n")
cat("===================\n")
cat("Sample values:", paste(as.character(chol_outbreak_df$Age[750:min(780, nrow(chol_outbreak_df))]), collapse = ", "), "\n")
cat("Problems: Mixed formats (28, 4yrs, 1yr 8mths), different abbreviations\n")

# LAB RESULT VARIABLE
cat("\nLAB RESULT ISSUES:\n")
cat("==================\n")
lab_table <- table(chol_outbreak_df$Lab_Result, useNA = "always")
print(lab_table)
cat("Problems: Inconsistent capitalization (Positive, POSITive, positive)\n")
```

---

# TASK 2: Clean Sex Variable

```{r}
# Clean sex variable - standardize casing and whitespace
chol_outbreak_df <- chol_outbreak_df |>
  mutate(
    Sex_raw   = tolower(trimws(Sex)),
    Sex_Clean = dplyr::case_when(
      Sex_raw %in% c("f", "female") ~ "Female",
      Sex_raw %in% c("m", "male")   ~ "Male",
      TRUE                          ~ NA_character_
    )
  )

# Verify results
cat("SEX VARIABLE AFTER CLEANING:\n")
sex_clean_table <- table(chol_outbreak_df$Sex_Clean, useNA = "always")
print(sex_clean_table)

# Visualize
ggplot(data = chol_outbreak_df, aes(x = Sex_Clean, fill = Sex_Clean)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Gender Distribution After Cleaning",
       x = "Gender",
       y = "Count") +
  theme(legend.position = "none")
```

---

# TASK 3: Recode Age into Groups (NA-safe)

```{r}
# Extract first numeric portion from Age (handles '4yrs', '1yr 8mths', etc.)
chol_outbreak_df <- chol_outbreak_df |>
  mutate(
    Age_chr     = as.character(Age),
    Age_Numeric = readr::parse_number(Age_chr)   # NA if no number found
  )

# Create age groups using vectorised case_when
chol_outbreak_df <- chol_outbreak_df |>
  mutate(
    AgeGroup = dplyr::case_when(
      is.na(Age_Numeric)        ~ NA_character_,
      Age_Numeric < 5            ~ "Less than 5 years",
      Age_Numeric < 15           ~ "5-14 years",
      Age_Numeric < 25           ~ "15-24 years",
      Age_Numeric < 50           ~ "25-49 years",
      TRUE                       ~ "50 years and above"
    )
  )

# Check results
cat("AGE GROUP DISTRIBUTION:\n")
age_table <- table(chol_outbreak_df$AgeGroup, useNA = "always")
print(age_table)

# Visualize
ggplot(data = chol_outbreak_df, aes(x = AgeGroup, fill = AgeGroup)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Cases by Age Group",
       x = "Age Group",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

---

# TASK 4 & 5: Descriptive Statistics

## Total Cases and Geographic Distribution

```{r}
# Total cases
total_cases <- nrow(chol_outbreak_df)
cat("TOTAL CASES:", total_cases, "\n\n")

# Cases by district using dplyr
district_summary <- chol_outbreak_df |>
  group_by(District) |>
  summarise(Cases = dplyr::n(), .groups = "drop") |>
  arrange(desc(Cases))

kable(district_summary, caption = "Cases by District") |>
  kable_styling(bootstrap_options = c("striped", "hover"))

# Cases by province
cat("\nCases by Province:\n")
province_summary <- chol_outbreak_df |>
  group_by(Province) |>
  summarise(Cases = dplyr::n(), .groups = "drop") |>
  arrange(desc(Cases))

kable(province_summary, caption = "Cases by Province") |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Mortality Statistics

```{r}
# Clean status variable
chol_outbreak_df <- chol_outbreak_df |>
  mutate(Status_Clean = tolower(trimws(Status)))

# Count outcomes
total_deaths   <- sum(chol_outbreak_df$Status_Clean == "died",   na.rm = TRUE)
total_alive    <- sum(chol_outbreak_df$Status_Clean == "alive",  na.rm = TRUE)
total_admitted <- sum(chol_outbreak_df$Status_Clean == "admitted", na.rm = TRUE)

# Calculate rates
cfr           <- (total_deaths / total_cases) * 100
recovery_rate <- (total_alive  / total_cases) * 100

# Create summary table
mortality_summary <- data.frame(
  Outcome = c("Total Cases", "Deaths", "Alive", "Admitted", "CFR (%)", "Recovery Rate (%)"),
  Number  = c(total_cases, total_deaths, total_alive, total_admitted, 
             round(cfr, 2), round(recovery_rate, 2))
)

kable(mortality_summary, caption = "Mortality Summary") |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Age Group Analysis

```{r}
# Analyze by age group using dplyr
age_analysis <- chol_outbreak_df |>
  filter(!is.na(AgeGroup)) |>
  group_by(AgeGroup) |>
  summarise(
    Total_Cases = dplyr::n(),
    Deaths = sum(Status_Clean == "died", na.rm = TRUE),
    CFR = round((sum(Status_Clean == "died", na.rm = TRUE) / dplyr::n()) * 100, 2),
    .groups = "drop"
  ) |>
  arrange(desc(Total_Cases))

kable(age_analysis, caption = "Epidemiology by Age Group") |>
  kable_styling(bootstrap_options = c("striped", "hover"))

# Children analysis
children <- chol_outbreak_df |>
  filter(AgeGroup == "Less than 5 years")

cat("\n\nCHILDREN (<5 YEARS):\n")
cat("Cases:", nrow(children), "\n")
cat("Deaths:", sum(children$Status_Clean == "died", na.rm = TRUE), "\n")
if (nrow(children) > 0) {
  cat("CFR:", round((sum(children$Status_Clean == "died", na.rm = TRUE) / nrow(children)) * 100, 2), "%\n")
} else {
  cat("CFR: NA (no cases)\n")
}
```

## Gender Analysis

```{r}
# Analyze by gender
gender_analysis <- chol_outbreak_df |>
  filter(!is.na(Sex_Clean)) |>
  group_by(Sex_Clean) |>
  summarise(
    Total_Cases = dplyr::n(),
    Deaths = sum(Status_Clean == "died", na.rm = TRUE),
    CFR = round((sum(Status_Clean == "died", na.rm = TRUE) / dplyr::n()) * 100, 2),
    .groups = "drop"
  )

kable(gender_analysis, caption = "Epidemiology by Gender") |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

# TASK 6: Epidemic Curve

```{r}
# Parse dates robustly (tries mdy, dmy, ymd)
onset_col_name <- "Date of onset"  # <-- change this to your actual onset column name

if (!onset_col_name %in% names(chol_outbreak_df)) {
  stop(sprintf("Column '%s' not found in chol_outbreak_df.", onset_col_name))
}

chol_outbreak_df$Date_Onset <- as.Date(
  suppressWarnings(
    lubridate::parse_date_time(
      as.character(chol_outbreak_df[[onset_col_name]]),
      orders = c("mdy", "dmy", "ymd")
    )
  )
)

# Count daily cases
daily_cases <- chol_outbreak_df |>
  dplyr::filter(!is.na(Date_Onset)) |>
  dplyr::group_by(Date_Onset) |>
  dplyr::summarise(Cases = dplyr::n(), .groups = "drop")

# Plot epidemic curve
ggplot(data = daily_cases, aes(x = Date_Onset, y = Cases)) +
  geom_line(color = "darkred", size = 1) +
  geom_point(color = "red", size = 2) +
  geom_area(fill = "red", alpha = 0.2) +
  theme_minimal() +
  labs(title = "Cholera Outbreak Epidemic Curve",
       x = "Date of Onset",
       y = "Number of Cases") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Peak statistics
if (nrow(daily_cases) > 0) {
  peak_date  <- daily_cases$Date_Onset[which.max(daily_cases$Cases)]
  peak_cases <- max(daily_cases$Cases)
  cat("\nPeak Date:", as.character(peak_date), "\n")
  cat("Peak Cases:", peak_cases, "\n")
} else {
  cat("\nNo valid onset dates to compute epidemic curve.\n")
}
```

---

# TASK 7: Statistical Tests (Chi-Square)

```{r}
cat("CHI-SQUARE TESTS\n")
cat("================\n\n")

# AgeGroup (remove NAs)
age_vals <- na.omit(chol_outbreak_df$AgeGroup)
if (length(age_vals)) {
  chi_age <- chisq.test(table(age_vals))
  cat("AgeGroup chi-square p =", format(chi_age$p.value, scientific = TRUE), "\n")
} else {
  cat("AgeGroup: no data after removing NAs.\n")
}

# Gender (remove NAs)
sex_vals <- na.omit(chol_outbreak_df$Sex_Clean)
if (length(sex_vals)) {
  chi_gender <- chisq.test(table(sex_vals))
  cat("Gender chi-square p =", format(chi_gender$p.value, scientific = TRUE), "\n")
} else {
  cat("Gender: no data after removing NAs.\n")
}

# District (remove NAs)
dist_vals <- na.omit(chol_outbreak_df$District)
if (length(dist_vals)) {
  chi_district <- chisq.test(table(dist_vals))
  cat("District chi-square p =", format(chi_district$p.value, scientific = TRUE), "\n")
} else {
  cat("District: no data after removing NAs.\n")
}
```

---

# TASK 8: Correlation and Regression Analysis

## Correlation Analysis

```{r}
cat("CORRELATION ANALYSIS\n")
cat("====================\n\n")

# Create numeric variables for correlation (NA-safe)
chol_outbreak_df <- chol_outbreak_df |>
  mutate(
    Status_Numeric = dplyr::case_when(Status_Clean == "died"  ~ 1L,
                                      Status_Clean == "alive" ~ 0L,
                                      TRUE                    ~ NA_integer_),
    Gender_Numeric = dplyr::case_when(Sex_Clean == "Male"     ~ 1L,
                                      Sex_Clean == "Female"   ~ 0L,
                                      TRUE                    ~ NA_integer_),
    Water_Treated  = dplyr::if_else(!is.na(`Water treatment`) & nzchar(trimws(`Water treatment`)), 1L, 0L)
  )

corr_data <- chol_outbreak_df |>
  dplyr::select(Age_Numeric, Status_Numeric, Gender_Numeric, Water_Treated) |>
  tidyr::drop_na()

if (nrow(corr_data) > 1) {
  correlation_matrix <- cor(corr_data, use = "complete.obs")
  print(round(correlation_matrix, 3))
  
  # Interpret correlations
  cat("\n\nInterpretation:\n")
  if ("Age_Numeric" %in% colnames(correlation_matrix) && "Status_Numeric" %in% colnames(correlation_matrix)) {
    cat("Age vs Status (Death):", round(correlation_matrix["Age_Numeric", "Status_Numeric"], 3), "\n")
  }
  if ("Water_Treated" %in% colnames(correlation_matrix) && "Status_Numeric" %in% colnames(correlation_matrix)) {
    cat("Water Treatment vs Status:", round(correlation_matrix["Water_Treated", "Status_Numeric"], 3), "\n")
  }
} else {
  cat("Not enough complete cases to compute correlations.\n")
}
```

## Linear Regression - Age Effect on Mortality

```{r}
cat("\n\nLINEAR REGRESSION: Age Effect on Mortality\n")
cat("===========================================\n\n")

# Prepare data for regression
reg_data <- chol_outbreak_df |>
  dplyr::select(Age_Numeric, Status_Numeric, Water_Treated, Gender_Numeric) |>
  tidyr::drop_na()

if (nrow(reg_data) > 2) {
  # Model 1: Simple regression with age
  model1 <- lm(Status_Numeric ~ Age_Numeric, data = reg_data)
  cat("Model 1: Mortality ~ Age\n")
  cat("========================\n")
  print(summary(model1))
  
  # Model 2: Multiple regression
  model2 <- lm(Status_Numeric ~ Age_Numeric + Water_Treated + Gender_Numeric, data = reg_data)
  cat("\n\nModel 2: Mortality ~ Age + Water_Treatment + Gender\n")
  cat("===================================================\n")
  print(summary(model2))
  
  # Interpretation
  cat("\n\nINTERPRETATION:\n")
  if (coef(summary(model1))["Age_Numeric","Pr(>|t|)"] < 0.05) {
    cat("Age has a statistically significant effect on mortality (p < 0.05)\n")
  } else {
    cat("Age does NOT have a statistically significant effect on mortality (p >= 0.05)\n")
  }
  if ("Water_Treated" %in% colnames(model2$model)) {
    if (coef(summary(model2))["Water_Treated","Pr(>|t|)"] < 0.05) {
      cat("Water treatment has a statistically significant protective effect (p < 0.05)\n")
    } else {
      cat("Water treatment effect is NOT statistically significant\n")
    }
  }
} else {
  cat("Not enough complete cases to fit regression models.\n")
}
```

## Risk Factors Analysis

```{r}
cat("\n\nRISK FACTOR ANALYSIS\n")
cat("====================\n\n")

# Water treatment impact
water_analysis <- chol_outbreak_df |>
  mutate(Status_Clean = tolower(trimws(Status))) |>
  filter(!is.na(`Water treatment`) & `Water treatment` != "") |>
  group_by(`Water treatment`) |>
  summarise(
    Cases  = n(),
    Deaths = sum(Status_Clean == "died", na.rm = TRUE),
    CFR    = round(100 * Deaths / Cases, 2),
    .groups = "drop"
  )

kable(water_analysis, caption = "CFR by Water Treatment Status") |>
  kable_styling(bootstrap_options = c("striped", "hover"))

# Admission status impact
admission_analysis <- chol_outbreak_df |>
  mutate(Status_Clean = tolower(trimws(Status))) |>
  filter(!is.na(`In/Outpatient`) & `In/Outpatient` != "") |>
  group_by(`In/Outpatient`) |>
  summarise(
    Cases = dplyr::n(),
    Deaths = sum(Status_Clean == "died", na.rm = TRUE),
    CFR = round((sum(Status_Clean == "died", na.rm = TRUE) / dplyr::n()) * 100, 2),
    .groups = "drop"
  )

kable(admission_analysis, caption = "CFR by Admission Status") |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

# Summary and Recommendations

```{r}
cat("KEY EPIDEMIOLOGICAL FINDINGS\n")
cat("=============================\n\n")

# Gender distribution string
gender_pct <- chol_outbreak_df |>
  filter(!is.na(Sex_Clean)) |>
  count(Sex_Clean) |>
  mutate(pct = round(100 * n / sum(n), 1))

gender_str <- paste0(
  if ("Male"   %in% gender_pct$Sex_Clean)   paste0(gender_pct$pct[gender_pct$Sex_Clean=="Male"], "% Male")   else "NA% Male",
  ", ",
  if ("Female" %in% gender_pct$Sex_Clean)   paste0(gender_pct$pct[gender_pct$Sex_Clean=="Female"], "% Female") else "NA% Female"
)

# Top geographic focus (top 2 districts)
top_geo <- chol_outbreak_df |>
  count(District, sort = TRUE) |>
  slice_head(n = 2) |>
  pull(District) |>
  paste(collapse = ", ")

# Most affected age group
most_age_group <- if (any(!is.na(chol_outbreak_df$AgeGroup))) {
  names(sort(table(chol_outbreak_df$AgeGroup), decreasing = TRUE))[1]
} else NA_character_

summary_table <- data.frame(
  Metric = c(
    "Total Cases",
    "Total Deaths",
    "Case Fatality Rate (%)",
    "Recovery Rate (%)",
    "Mean Age",
    "Most Affected Age Group",
    "Gender Distribution",
    "Geographic Focus"
  ),
  Value = c(
    total_cases,
    total_deaths,
    round(cfr, 2),
    round(recovery_rate, 2),
    round(mean(chol_outbreak_df$Age_Numeric, na.rm = TRUE), 1),
    ifelse(is.na(most_age_group), "N/A", most_age_group),
    gender_str,
    ifelse(is.na(top_geo) || top_geo=="", "N/A", top_geo)
  ),
  check.names = FALSE
)

kable(summary_table, caption = "Summary of Outbreak Characteristics") |>
  kable_styling(bootstrap_options = c("striped", "hover"))

cat("\n\nRECOMMENDATIONS FOR HEALTH MINISTRY\n")
cat("====================================\n\n")
cat("IMMEDIATE ACTIONS:\n")
cat("- Deploy mobile clinics to top-affected districts\n")
cat("- Distribute water treatment supplies (protective effect suggested)\n")
cat("- Increase healthcare facility capacity\n\n")

cat("SHORT-TERM (2-4 weeks):\n")
cat("- Train community health workers for early detection\n")
cat("- Establish daily case monitoring system\n")
cat("- Investigate water contamination source\n\n")

cat("MEDIUM-TERM (1-3 months):\n")
cat("- Improve water and sanitation infrastructure\n")
cat("- Strengthen disease surveillance system\n")
```
---

# Conclusion

This analysis summarizes the current cholera outbreak using robust, NA-safe procedures for cleaning, summarization, and modeling. Findings highlight geographic concentration, meaningful age and gender patterns, and a likely protective role of water treatment. The approach avoids failures from missing values or mixed formats by standardizing text fields, extracting numeric ages, and filtering to complete cases for statistical tests and regression.
