---
title: "Cholera Outbreak Analysis Report"
subtitle: "Statistical Analysis for Ministry of Health Briefing"
author: "Thierry SEBAKUNZI | ID:101394"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: bootstrap
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 10, fig.height = 6)
```

# Executive Summary

This report analyzes the cholera outbreak data from two provinces to provide the Minister of Health with comprehensive insights for resource allocation and response planning.

# 1. Data Loading and Initial Examination

```{r load-libraries}
# Load required libraries
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)
library(kableExtra)
library(lubridate)
library(knitr)
library(DT)
library(plotly)
library(gridExtra)
library(scales)
```

```{r load-data}
# Load the data
data <- read_excel("DataFile.xls")

# Display basic information about the dataset
cat("Dataset dimensions:", dim(data)[1], "rows,", dim(data)[2], "columns\n")
cat("Column names:", paste(names(data), collapse = ", "), "\n")

# Display first few rows
head(data) %>% kable(caption = "First 6 rows of the dataset")
```

```{r data-structure}
# Examine data structure
str(data)
summary(data)
```

# 2. Data Quality Assessment

```{r data-quality}
# Check for missing values
missing_summary <- data %>%
  summarise_all(~sum(is.na(.))) %>%
  gather(Variable, Missing_Count) %>%
  mutate(Missing_Percentage = round(Missing_Count / nrow(data) * 100, 2))

missing_summary %>% 
  kable(caption = "Missing Values Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Check unique values in key variables
cat("\nUnique values in Sex variable:\n")
table(data$Sex, useNA = "ifany")

cat("\nUnique values in Outcome variable:\n")
table(data$Outcome, useNA = "ifany")

cat("\nUnique values in District variable:\n")
table(data$District, useNA = "ifany")

# Check age distribution
cat("\nAge variable summary:\n")
summary(data$Age)
```

## Data Quality Issues Identified:

1. **Missing Values**: Check for any missing data in critical variables
2. **Sex Variable Inconsistency**: Multiple formats for gender coding
3. **Date Format**: Ensure date variables are properly formatted
4. **Age Outliers**: Check for unrealistic age values
5. **Outcome Coding**: Verify consistent outcome coding

# 3. Data Cleaning and Harmonization

```{r harmonize-sex}
# Harmonize Sex variable
data <- data %>%
  mutate(Sex_Clean = case_when(
    tolower(Sex) %in% c("m", "male", "homme") ~ "Male",
    tolower(Sex) %in% c("f", "female", "femme") ~ "Female",
    TRUE ~ "Unknown"
  ))

# Check the harmonization
table(data$Sex, data$Sex_Clean, useNA = "ifany")
```

```{r recode-age-groups}
# Recode Age into AgeGroup
data <- data %>%
  mutate(AgeGroup = case_when(
    Age < 5 ~ "Less than 5 years",
    Age >= 5 & Age <= 14 ~ "5-14 years",
    Age >= 15 & Age <= 24 ~ "15-24 years",
    Age >= 25 & Age <= 49 ~ "25-49 years",
    Age >= 50 ~ "50 years and above",
    TRUE ~ "Unknown"
  )) %>%
  mutate(AgeGroup = factor(AgeGroup, levels = c("Less than 5 years", "5-14 years", 
                                                "15-24 years", "25-49 years", 
                                                "50 years and above", "Unknown")))

# Check age group distribution
table(data$AgeGroup, useNA = "ifany")
```

```{r date-processing}
# Process date variables (assuming DateOnset exists)
if("DateOnset" %in% names(data)) {
  data$DateOnset <- as.Date(data$DateOnset)
  data$EpiWeek <- epiweek(data$DateOnset)
}
```

# 4. Statistical Analysis for Minister's Briefing

## 4.1 Total Number of Cases and Geographical Distribution

```{r total-cases}
total_cases <- nrow(data)
cat("Total number of cases:", total_cases, "\n")

# Geographical distribution
geo_dist <- data %>%
  group_by(District) %>%
  summarise(
    Cases = n(),
    Percentage = round(n()/nrow(data)*100, 1)
  ) %>%
  arrange(desc(Cases))

geo_dist %>% 
  kable(caption = "Geographical Distribution of Cases") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## 4.2 Deaths and Case Fatality Rate

```{r deaths-cfr}
# Calculate deaths and CFR
outcome_summary <- data %>%
  group_by(Outcome) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = round(Count/sum(Count)*100, 1))

deaths <- sum(data$Outcome == "Death" | data$Outcome == "Died", na.rm = TRUE)
cfr <- round(deaths/total_cases*100, 2)

cat("Number of deaths:", deaths, "\n")
cat("Case Fatality Rate:", cfr, "%\n")

outcome_summary %>% 
  kable(caption = "Distribution by Outcome") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## 4.3 Age Group Analysis

```{r age-analysis}
# Age group distribution
age_dist <- data %>%
  group_by(AgeGroup) %>%
  summarise(
    Cases = n(),
    Deaths = sum(Outcome %in% c("Death", "Died"), na.rm = TRUE),
    CFR = round(Deaths/Cases*100, 1),
    Percentage = round(Cases/nrow(data)*100, 1)
  ) %>%
  arrange(desc(Cases))

age_dist %>% 
  kable(caption = "Cases and Deaths by Age Group") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Highlight children (<5 years)
children_cases <- age_dist$Cases[age_dist$AgeGroup == "Less than 5 years"]
children_percentage <- age_dist$Percentage[age_dist$AgeGroup == "Less than 5 years"]

cat("\nChildren under 5 years:\n")
cat("Cases:", children_cases, "(", children_percentage, "% of total cases)\n")
```

## 4.4 Gender Distribution

```{r gender-analysis}
# Gender distribution
gender_dist <- data %>%
  group_by(Sex_Clean) %>%
  summarise(
    Cases = n(),
    Deaths = sum(Outcome %in% c("Death", "Died"), na.rm = TRUE),
    CFR = round(Deaths/Cases*100, 1),
    Percentage = round(Cases/nrow(data)*100, 1)
  )

gender_dist %>% 
  kable(caption = "Cases and Deaths by Gender") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## 4.5 Statistical Tests for Equality Across Groups

```{r statistical-tests}
# Chi-square test for age groups
age_chi <- chisq.test(table(data$AgeGroup))
cat("Chi-square test for equal distribution across age groups:\n")
cat("Chi-square statistic:", round(age_chi$statistic, 3), "\n")
cat("p-value:", format(age_chi$p.value, scientific = TRUE), "\n")

# Chi-square test for gender
gender_chi <- chisq.test(table(data$Sex_Clean))
cat("\nChi-square test for equal distribution across genders:\n")
cat("Chi-square statistic:", round(gender_chi$statistic, 3), "\n")
cat("p-value:", format(gender_chi$p.value, scientific = TRUE), "\n")

# Chi-square test for districts
district_chi <- chisq.test(table(data$District))
cat("\nChi-square test for equal distribution across districts:\n")
cat("Chi-square statistic:", round(district_chi$statistic, 3), "\n")
cat("p-value:", format(district_chi$p.value, scientific = TRUE), "\n")
```

## 4.6 Recovery Rate Analysis

```{r recovery-rate}
# Calculate recovery rate
recovery_summary <- data %>%
  group_by(Outcome) %>%
  summarise(Count = n()) %>%
  mutate(
    Percentage = round(Count/sum(Count)*100, 1),
    Rate = case_when(
      Outcome %in% c("Recovered", "Recovery", "Alive") ~ "Recovery Rate",
      Outcome %in% c("Death", "Died", "Dead") ~ "Death Rate",
      TRUE ~ "Other"
    )
  )

recovered <- sum(data$Outcome %in% c("Recovered", "Recovery", "Alive"), na.rm = TRUE)
recovery_rate <- round(recovered/total_cases*100, 2)

cat("Recovery rate:", recovery_rate, "%\n")

recovery_summary %>% 
  kable(caption = "Outcome Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# 5. Visualizations

## 5.1 Epidemic Curve

```{r epidemic-curve}
if("DateOnset" %in% names(data) && !all(is.na(data$DateOnset))) {
  # Create epidemic curve
  epi_curve_data <- data %>%
    filter(!is.na(DateOnset)) %>%
    group_by(DateOnset) %>%
    summarise(Cases = n()) %>%
    arrange(DateOnset)
  
  p1 <- ggplot(epi_curve_data, aes(x = DateOnset, y = Cases)) +
    geom_col(fill = "steelblue", alpha = 0.7) +
    geom_smooth(method = "loess", se = FALSE, color = "red") +
    labs(title = "Epidemic Curve - Cholera Outbreak",
         subtitle = "Daily Number of Cases by Date of Onset",
         x = "Date of Onset",
         y = "Number of Cases") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  print(p1)
} else {
  cat("Date of onset data not available for epidemic curve\n")
}
```

## 5.2 Geographical Distribution

```{r geo-visualization}
# Bar chart of cases by district
p2 <- ggplot(geo_dist, aes(x = reorder(District, Cases), y = Cases)) +
  geom_col(fill = "coral", alpha = 0.8) +
  geom_text(aes(label = Cases), hjust = -0.1) +
  coord_flip() +
  labs(title = "Geographical Distribution of Cholera Cases",
       x = "District",
       y = "Number of Cases") +
  theme_minimal()

print(p2)
```

## 5.3 Age Group Distribution

```{r age-visualization}
# Bar chart of cases by age group
p3 <- ggplot(age_dist, aes(x = AgeGroup, y = Cases)) +
  geom_col(fill = "lightgreen", alpha = 0.8) +
  geom_text(aes(label = Cases), vjust = -0.5) +
  labs(title = "Distribution of Cholera Cases by Age Group",
       x = "Age Group",
       y = "Number of Cases") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p3)

# CFR by age group
p4 <- ggplot(age_dist, aes(x = AgeGroup, y = CFR)) +
  geom_col(fill = "orange", alpha = 0.8) +
  geom_text(aes(label = paste0(CFR, "%")), vjust = -0.5) +
  labs(title = "Case Fatality Rate by Age Group",
       x = "Age Group",
       y = "Case Fatality Rate (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p4)
```

## 5.4 Gender Distribution

```{r gender-visualization}
# Pie chart for gender distribution
p5 <- ggplot(gender_dist, aes(x = "", y = Cases, fill = Sex_Clean)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(title = "Gender Distribution of Cholera Cases",
       fill = "Gender") +
  theme_void() +
  geom_text(aes(label = paste0(Cases, "\n(", Percentage, "%)")), 
            position = position_stack(vjust = 0.5))

print(p5)
```

## 5.5 Outcome Distribution

```{r outcome-visualization}
# Bar chart of outcomes
p6 <- ggplot(outcome_summary, aes(x = Outcome, y = Count)) +
  geom_col(fill = "purple", alpha = 0.7) +
  geom_text(aes(label = paste0(Count, "\n(", Percentage, "%)")), vjust = -0.5) +
  labs(title = "Distribution of Case Outcomes",
       x = "Outcome",
       y = "Number of Cases") +
  theme_minimal()

print(p6)
```

# 6. Risk Factor Analysis

```{r risk-factors}
# Logistic regression for death outcome
if(sum(data$Outcome %in% c("Death", "Died"), na.rm = TRUE) > 0) {
  
  # Create binary death outcome
  data$Death <- ifelse(data$Outcome %in% c("Death", "Died"), 1, 0)
  
  # Logistic regression model
  model <- glm(Death ~ AgeGroup + Sex_Clean + District, 
               data = data, family = binomial)
  
  cat("Logistic Regression Results for Death Outcome:\n")
  summary(model)
  
  # Odds ratios
  or_results <- exp(coef(model))
  cat("\nOdds Ratios:\n")
  print(round(or_results, 3))
  
} else {
  cat("No deaths recorded for risk factor analysis\n")
}
```

# 7. Key Findings Summary

## For the Minister of Health:

1. **Total Cases**: `r total_cases` cholera cases recorded across two provinces

2. **Geographical Impact**: 
   - Most affected district: `r geo_dist$District[1]` (`r geo_dist$Cases[1]` cases)
   - Least affected district: `r geo_dist$District[nrow(geo_dist)]` (`r geo_dist$Cases[nrow(geo_dist)]` cases)

3. **Mortality**: 
   - Deaths: `r deaths`
   - Case Fatality Rate: `r cfr`%

4. **Age Groups**:
   - Most affected: `r age_dist$AgeGroup[1]` (`r age_dist$Cases[1]` cases)
   - Children under 5: `r children_cases` cases (`r children_percentage`% of total)

5. **Gender Distribution**: 
   - Males: `r gender_dist$Cases[gender_dist$Sex_Clean == "Male"]` cases
   - Females: `r gender_dist$Cases[gender_dist$Sex_Clean == "Female"]` cases

6. **Recovery Rate**: `r recovery_rate`%

# 8. Recommendations for Surveillance and Outbreak Response

Based on the analysis, I recommend:

## Immediate Actions:
1. **Resource Allocation**: Focus resources on the most affected districts
2. **Targeted Interventions**: Prioritize age groups with highest case fatality rates
3. **Enhanced Surveillance**: Strengthen case detection in underreporting areas

## Data Quality Improvements:
1. **Standardization**: Implement standardized data collection forms
2. **Training**: Train data collectors on consistent variable coding
3. **Validation**: Implement real-time data validation checks
4. **Completeness**: Ensure all critical variables are captured

## Surveillance Strengthening:
1. **Early Warning**: Establish early warning systems in high-risk areas
2. **Community Engagement**: Involve communities in case identification
3. **Laboratory Capacity**: Strengthen laboratory confirmation capabilities
4. **Contact Tracing**: Implement systematic contact tracing

## Response Optimization:
1. **Treatment Centers**: Establish treatment centers in affected areas
2. **WASH Interventions**: Improve water, sanitation, and hygiene
3. **Health Education**: Conduct community health education campaigns
4. **Supply Chain**: Ensure adequate medical supplies and ORS

---

*Analysis completed on `r Sys.Date()` using R version `r R.version.string`*