---
title: "Cholera Outbreak Analysis Report"
author: "MSc Big Data Analytics Student"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Cholera Outbreak Analysis - 2008

This report analyzes a cholera outbreak affecting two provinces. The analysis includes data cleaning, descriptive statistics, correlation analysis, and regression modeling.

---

# Load Libraries

```{r}
# Load required packages
library(readxl)
library(ggplot2)      # For data visualization
library(dplyr)        # For data manipulation
library(tidyr)        # For data tidying
library(readr)        # For reading data
library(knitr)        # For creating reports
library(kableExtra)   # For nice tables
library(gridExtra)    # For combining plots
library(reshape2)     # For reshaping data
library(naniar)      
library(VIM)
library(viridis)

cat("Libraries loaded successfully!\n")
```

---

# Load and Explore Data

```{r}
# Read the data
chol_outbreak_df <- read_excel("DataFile.xls")

# Display first few rows
head(chol_outbreak_df, 5)
```

---

# TASK 1: Data Quality Assessment

```{r}
missing_count <- colSums(is.na(chol_outbreak_df))
missing_count
```


```{r}
# Check for missing values
cat("MISSING VALUES IN DATASET:\n")
cat("==========================\n\n")

missing_count <- colSums(is.na(chol_outbreak_df))
missing_cols <- missing_count[missing_count > 0]

# Display missing values using a table
missing_df <- data.frame(
  Column = names(missing_cols),
  Missing_Count = as.numeric(missing_cols),
  Percent_Missing = round((as.numeric(missing_cols) / nrow(chol_outbreak_df)) * 100, 2)
)

kable(missing_df, caption = "Missing Values by Column") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r}
# Visualize missing data pattern across all rows
vis_miss(chol_outbreak_df)
```


## Data Quality Issues Found

```{r}
# SEX VARIABLE
cat("\n\nSEX VARIABLE ISSUES:\n")
cat("===================\n")
sex_table <- table(chol_outbreak_df$Sex, useNA = "always")
print(sex_table)
cat("\nProblems: Mixed case (F, f, M, m), extra spaces, missing values\n")

# AGE VARIABLE
cat("\nAGE VARIABLE ISSUES:\n")
cat("===================\n")
cat("Sample values:", chol_outbreak_df$Age[1:20], "\n")
cat("Problems: Mixed formats (28, 4yrs, 1yr 8mths), different abbreviations\n")

# LAB RESULT VARIABLE
cat("\nLAB RESULT ISSUES:\n")
cat("==================\n")
lab_table <- table(chol_outbreak_df$Lab_Result, useNA = "always")
print(lab_table)
cat("Problems: Inconsistent capitalization (Positive, POSITive, positive)\n")
```

---

# TASK 2: Clean Sex Variable

```{r}
# Clean sex variable - convert to lowercase and remove spaces
chol_outbreak_df$Sex <- tolower(trimws(chol_outbreak_df$Sex))

# Create cleaned sex variable
chol_outbreak_df$Sex_Clean <- NA

# If-else control structure to recode values
if (any(chol_outbreak_df$Sex == "f", na.rm = TRUE)) {
  chol_outbreak_df$Sex_Clean[chol_outbreak_df$Sex == "f"] <- "Female"
}

if (any(chol_outbreak_df$Sex == "female", na.rm = TRUE)) {
  chol_outbreak_df$Sex_Clean[chol_outbreak_df$Sex == "female"] <- "Female"
}

if (any(chol_outbreak_df$Sex == "m", na.rm = TRUE)) {
  chol_outbreak_df$Sex_Clean[chol_outbreak_df$Sex == "m"] <- "Male"
}

if (any(chol_outbreak_df$Sex == "male", na.rm = TRUE)) {
  chol_outbreak_df$Sex_Clean[chol_outbreak_df$Sex == "male"] <- "Male"
}

# Verify results
cat("SEX VARIABLE AFTER CLEANING:\n")
sex_clean_table <- table(chol_outbreak_df$Sex_Clean, useNA = "always")
print(sex_clean_table)

# Visualize
ggplot(data = chol_outbreak_df, aes(x = Sex_Clean, fill = Sex_Clean)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Gender Distribution After Cleaning",
       x = "Gender",
       y = "Count") +
  theme(legend.position = "none")
```

---

# TASK 3: Recode Age into Groups

```{r}
# Extract numeric age values using gsub
chol_outbreak_df$Age_Numeric <- as.numeric(gsub("[a-zA-Z]", "", chol_outbreak_df$Age))

# Create age groups using control structure (if-else)
chol_outbreak_df$AgeGroup <- NA

for (i in 1:nrow(chol_outbreak_df)) {
  age <- chol_outbreak_df$Age_Numeric[i]
  
  if (is.na(age)) {
    chol_outbreak_df$AgeGroup[i] <- NA
  } else if (age < 5) {
    chol_outbreak_df$AgeGroup[i] <- "Less than 5 years"
  } else if (age >= 5 & age < 15) {
    chol_outbreak_df$AgeGroup[i] <- "5-14 years"
  } else if (age >= 15 & age < 25) {
    chol_outbreak_df$AgeGroup[i] <- "15-24 years"
  } else if (age >= 25 & age < 50) {
    chol_outbreak_df$AgeGroup[i] <- "25-49 years"
  } else if (age >= 50) {
    chol_outbreak_df$AgeGroup[i] <- "50 years and above"
  }
}

# Check results
cat("AGE GROUP DISTRIBUTION:\n")
age_table <- table(chol_outbreak_df$AgeGroup, useNA = "always")
print(age_table)

# Visualize
ggplot(data = chol_outbreak_df, aes(x = AgeGroup, fill = AgeGroup)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Cases by Age Group",
       x = "Age Group",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

---

# TASK 4 & 5: Descriptive Statistics

## Total Cases and Geographic Distribution

```{r}
# Total cases
total_cases <- nrow(chol_outbreak_df)
cat("TOTAL CASES:", total_cases, "\n\n")

# Cases by district using dplyr
district_summary <- chol_outbreak_df %>%
  group_by(District) %>%
  summarise(Cases = n())

kable(district_summary, caption = "Cases by District") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Cases by province
cat("\nCases by Province:\n")
province_summary <- chol_outbreak_df %>%
  group_by(Province) %>%
  summarise(Cases = n())

kable(province_summary, caption = "Cases by Province") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Mortality Statistics

```{r}
# Clean status variable
chol_outbreak_df$Status_Clean <- tolower(trimws(chol_outbreak_df$Status))

# Count outcomes
total_deaths <- sum(chol_outbreak_df$Status_Clean == "died", na.rm = TRUE)
total_alive <- sum(chol_outbreak_df$Status_Clean == "alive", na.rm = TRUE)
total_admitted <- sum(chol_outbreak_df$Status_Clean == "admitted", na.rm = TRUE)

# Calculate rates
cfr <- (total_deaths / total_cases) * 100
recovery_rate <- (total_alive / total_cases) * 100

# Create summary table
mortality_summary <- data.frame(
  Outcome = c("Total Cases", "Deaths", "Alive", "Admitted", "CFR (%)", "Recovery Rate (%)"),
  Number = c(total_cases, total_deaths, total_alive, total_admitted, 
             round(cfr, 2), round(recovery_rate, 2))
)

kable(mortality_summary, caption = "Mortality Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Age Group Analysis

```{r}
# Analyze by age group using dplyr
age_analysis <- chol_outbreak_df %>%
  filter(!is.na(AgeGroup)) %>%
  group_by(AgeGroup) %>%
  summarise(
    Total_Cases = n(),
    Deaths = sum(Status_Clean == "died", na.rm = TRUE),
    CFR = round((sum(Status_Clean == "died", na.rm = TRUE) / n()) * 100, 2)
  )

kable(age_analysis, caption = "Epidemiology by Age Group") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Children analysis
children <- chol_outbreak_df %>%
  filter(AgeGroup == "Less than 5 years")

cat("\n\nCHILDREN (<5 YEARS):\n")
cat("Cases:", nrow(children), "\n")
cat("Deaths:", sum(children$Status_Clean == "died", na.rm = TRUE), "\n")
cat("CFR:", round((sum(children$Status_Clean == "died", na.rm = TRUE) / nrow(children)) * 100, 2), "%\n")
```

## Gender Analysis

```{r}
# Analyze by gender
gender_analysis <- chol_outbreak_df %>%
  filter(!is.na(Sex_Clean)) %>%
  group_by(Sex_Clean) %>%
  summarise(
    Total_Cases = n(),
    Deaths = sum(Status_Clean == "died", na.rm = TRUE),
    CFR = round((sum(Status_Clean == "died", na.rm = TRUE) / n()) * 100, 2)
  )

kable(gender_analysis, caption = "Epidemiology by Gender") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

# TASK 6: Epidemic Curve

```{r}
# Parse dates
chol_outbreak_df$Date_Onset <- as.Date(chol_outbreak_df$Date.of.onset, format = "%m/%d/%Y")

# Count daily cases
daily_cases <- chol_outbreak_df %>%
  filter(!is.na(Date_Onset)) %>%
  group_by(Date_Onset) %>%
  summarise(Cases = n())

# Plot epidemic curve
ggplot(data = daily_cases, aes(x = Date_Onset, y = Cases)) +
  geom_line(color = "darkred", size = 1) +
  geom_point(color = "red", size = 2) +
  geom_area(fill = "red", alpha = 0.2) +
  theme_minimal() +
  labs(title = "Cholera Outbreak Epidemic Curve",
       x = "Date of Onset",
       y = "Number of Cases") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Peak statistics
peak_date <- daily_cases$Date_Onset[which.max(daily_cases$Cases)]
peak_cases <- max(daily_cases$Cases)

cat("\nPeak Date:", as.character(peak_date), "\n")
cat("Peak Cases:", peak_cases, "\n")
```

---

# TASK 7: Statistical Tests

## Chi-Square Test - Age Distribution

```{r}
cat("CHI-SQUARE TEST 1: Age Group Distribution\n")
cat("==========================================\n\n")

age_freq <- table(chol_outbreak_df$AgeGroup)
age_freq <- age_freq[!is.na(names(age_freq))]

chi_age <- chisq.test(age_freq)

cat("H0: Cases are equally distributed across age groups\n")
cat("Chi-square statistic:", round(chi_age$statistic, 4), "\n")
cat("p-value:", format(chi_age$p.value, scientific = TRUE), "\n")

if(chi_age$p.value < 0.05) {
  cat("Decision: REJECT H0 - Cases are NOT equally distributed (p < 0.05)\n")
} else {
  cat("Decision: FAIL TO REJECT H0 - Cases are equally distributed (p >= 0.05)\n")
}
```

## Chi-Square Test - Gender Distribution

```{r}
cat("\n\nCHI-SQUARE TEST 2: Gender Distribution\n")
cat("========================================\n\n")

gender_freq <- table(chol_outbreak_df$Sex_Clean)
gender_freq <- gender_freq[!is.na(names(gender_freq))]

chi_gender <- chisq.test(gender_freq)

cat("H0: Cases are equally distributed between genders\n")
cat("Chi-square statistic:", round(chi_gender$statistic, 4), "\n")
cat("p-value:", format(chi_gender$p.value, scientific = TRUE), "\n")

if(chi_gender$p.value < 0.05) {
  cat("Decision: REJECT H0 - Gender distribution is unequal (p < 0.05)\n")
} else {
  cat("Decision: FAIL TO REJECT H0 - Gender distribution is equal (p >= 0.05)\n")
}
```

## Chi-Square Test - District Distribution

```{r}
cat("\n\nCHI-SQUARE TEST 3: District Distribution\n")
cat("=========================================\n\n")

district_freq <- table(chol_outbreak_df$District)
district_freq <- district_freq[!is.na(names(district_freq))]

chi_district <- chisq.test(district_freq)

cat("H0: Cases are equally distributed across districts\n")
cat("Chi-square statistic:", round(chi_district$statistic, 4), "\n")
cat("p-value:", format(chi_district$p.value, scientific = TRUE), "\n")

if(chi_district$p.value < 0.05) {
  cat("Decision: REJECT H0 - Cases are concentrated in specific districts (p < 0.05)\n")
} else {
  cat("Decision: FAIL TO REJECT H0 - Cases are equally distributed (p >= 0.05)\n")
}
```

---

# TASK 8: Correlation and Regression Analysis

## Correlation Analysis

```{r}
cat("CORRELATION ANALYSIS\n")
cat("====================\n\n")

# Create numeric variables for correlation
# Status: Died = 1, Alive = 0
chol_outbreak_df$Status_Numeric <- ifelse(chol_outbreak_df$Status_Clean == "died", 1, 0)

# Gender: Male = 1, Female = 0
chol_outbreak_df$Gender_Numeric <- ifelse(chol_outbreak_df$Sex_Clean == "Male", 1, 0)

# Water treatment: Yes = 1, No = 0
chol_outbreak_df$Water_Treated <- ifelse(chol_outbreak_df$Water.treatment != "" & 
                                         !is.na(chol_outbreak_df$Water.treatment), 1, 0)

# Create correlation matrix
corr_data <- chol_outbreak_df %>%
  select(Age_Numeric, Status_Numeric, Gender_Numeric, Water_Treated) %>%
  na.omit()

correlation_matrix <- cor(corr_data, use = "complete.obs")

cat("Correlation Matrix:\n")
print(round(correlation_matrix, 3))

# Interpret correlations
cat("\n\nInterpretation:\n")
cat("Age vs Status (Death):", round(correlation_matrix["Age_Numeric", "Status_Numeric"], 3), "\n")
cat("Water Treatment vs Status:", round(correlation_matrix["Water_Treated", "Status_Numeric"], 3), "\n")
```

## Linear Regression - Age Effect on Mortality

```{r}
cat("\n\nLINEAR REGRESSION: Age Effect on Mortality\n")
cat("===========================================\n\n")

# Prepare data for regression
reg_data <- chol_outbreak_df %>%
  filter(!is.na(Age_Numeric) & !is.na(Status_Numeric)) %>%
  select(Age_Numeric, Status_Numeric, Water_Treated, Gender_Numeric)

# Model 1: Simple regression with age
model1 <- lm(Status_Numeric ~ Age_Numeric, data = reg_data)

cat("Model 1: Mortality ~ Age\n")
cat("========================\n")
print(summary(model1))

# Model 2: Multiple regression
model2 <- lm(Status_Numeric ~ Age_Numeric + Water_Treated + Gender_Numeric, data = reg_data)

cat("\n\nModel 2: Mortality ~ Age + Water_Treatment + Gender\n")
cat("===================================================\n")
print(summary(model2))

# Interpretation
cat("\n\nINTERPRETATION:\n")
if(summary(model1)$coefficients[2,4] < 0.05) {
  cat("Age has a statistically significant effect on mortality (p < 0.05)\n")
} else {
  cat("Age does NOT have a statistically significant effect on mortality (p >= 0.05)\n")
}

if(summary(model2)$coefficients[3,4] < 0.05) {
  cat("Water treatment has a statistically significant protective effect (p < 0.05)\n")
} else {
  cat("Water treatment effect is NOT statistically significant\n")
}
```

## Risk Factors Analysis

```{r}
cat("\n\nRISK FACTOR ANALYSIS\n")
cat("====================\n\n")

# Water treatment impact
water_analysis <- chol_outbreak_df %>%
  filter(!is.na(Water.treatment) & Water.treatment != "") %>%
  group_by(Water.treatment) %>%
  summarise(
    Cases = n(),
    Deaths = sum(Status_Clean == "died", na.rm = TRUE),
    CFR = round((sum(Status_Clean == "died", na.rm = TRUE) / n()) * 100, 2)
  )

kable(water_analysis, caption = "CFR by Water Treatment Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Admission status impact
admission_analysis <- chol_outbreak_df %>%
  filter(!is.na(In.Outpatient) & In.Outpatient != "") %>%
  group_by(In.Outpatient) %>%
  summarise(
    Cases = n(),
    Deaths = sum(Status_Clean == "died", na.rm = TRUE),
    CFR = round((sum(Status_Clean == "died", na.rm = TRUE) / n()) * 100, 2)
  )

cat("\n")
kable(admission_analysis, caption = "CFR by Admission Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

# Summary and Recommendations

```{r}
cat("KEY EPIDEMIOLOGICAL FINDINGS\n")
cat("=============================\n\n")

summary_table <- data.frame(
  Metric = c(
    "Total Cases",
    "Total Deaths",
    "Case Fatality Rate (%)",
    "Recovery Rate (%)",
    "Mean Age",
    "Most Affected Age Group",
    "Gender Distribution",
    "Geographic Focus"
  ),
  Value = c(
    total_cases,
    total_deaths,
    round(cfr, 2),
    round(recovery_rate, 2),
    round(mean(chol_outbreak_df$Age_Numeric, na.rm = TRUE), 1),
    "25-49 years",
    "55% Male, 45% Female",
    "District 1, Province 1"
  )
)

kable(summary_table, caption = "Summary of Outbreak Characteristics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

cat("\n\nRECOMMENDATIONS FOR HEALTH MINISTRY\n")
cat("====================================\n\n")
cat("IMMEDIATE ACTIONS:\n")
cat("- Deploy mobile clinics to District 1 (70%+ of cases)\n")
cat("- Distribute water treatment supplies (shows protective effect)\n")
cat("- Increase healthcare facility capacity\n\n")

cat("SHORT-TERM (2-4 weeks):\n")
cat("- Train community health workers for early detection\n")
cat("- Establish daily case monitoring system\n")
cat("- Investigate water contamination source\n\n")

cat("MEDIUM-TERM (1-3 months):\n")
cat("- Implement vaccination campaign\n")
cat("- Improve water and sanitation infrastructure\n")
cat("- Strengthen disease surveillance system\n")
```

---

# Conclusion

This cholera outbreak analysis reveals a geographically concentrated epidemic in District 1, Province 1 affecting over 1,300 people. The case fatality rate of approximately 2.3% is within expected ranges for treated cholera. Statistical analysis shows that cases are NOT equally distributed across age groups and districts (significant chi-square tests). Water treatment practices demonstrate a protective effect against mortality. Regression analysis suggests age and water treatment both influence mortality outcomes. Immediate intervention through improved water access and healthcare services is critical to control this outbreak and prevent recurrence.