---
title: "World Population Trend and CO2 Emission Analysis"
author: "Thierry SEBAKUNZI | Student ID: 101394"
date: "2025-10-10"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
# Quiet, consistent knitting
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Try to ensure UTF-8 during knit (won't change your data objects)
try(suppressWarnings(Sys.setlocale("LC_CTYPE", "C.UTF-8")), silent = TRUE)

# --- Safe table printer: DOES NOT modify your data ---
kable <- function(df, ...) {
  # Work on a copy
  x <- df
  # Fix header encoding + replace literal pipes that break kable
  cn <- iconv(names(x), from = "", to = "UTF-8", sub = ".")
  cn <- gsub("\\|", "/", cn, useBytes = TRUE)
  names(x) <- cn
  # Normalize only character columns for display
  x[] <- lapply(x, function(col) if (is.character(col)) iconv(col, from = "", to = "UTF-8", sub = ".") else col)
  knitr::kable(x, col.names = names(x), ...)
}

# --- Safe select: avoids crashing if a column is missing ---
pick_cols <- function(df, cols) dplyr::select(df, dplyr::any_of(cols))

```

# 1. Loading data to R studio

```{r libraries}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(corrplot)
library(plotly)
library(knitr)
library(DT)
library(reshape2)
library(viridis)
library(gt)

# Load datasets
population_data <- read.csv("world_population.csv")
co2_data <- read.csv("CO2_emission.csv")

print("Data loaded successfully!")
```

# 2. Exploratory Data Analysis

## 2.1 World Population Dataset Overview

```{r eda_population}
# Displaying variable names
cat("Variable names in Population Dataset:\n")
names(population_data)

```

```{r}
# Displaying top 5 rows
cat("\nTop 5 rows:\n")
head(population_data, 5)%>% kable()
```

```{r}
# Displaying bottom 10 rows
cat("\nBottom 10 rows:\n")
tail(population_data, 10) %>% kable()
```


```{r}
# Checking Data types
cat("\nData types:\n")
str(population_data)

# Checking Dataset shape
cat("\nDataset shape:\n")
cat("Rows:", nrow(population_data), "Columns:", ncol(population_data))
```


## 2.2 Dealing with Duplicates and Missing values

```{r data_quality}
# Check for duplicates
duplicates <- sum(duplicated(population_data))
cat("Number of duplicate rows:", duplicates)
```

```{r}
# Checking for missing values and displaying table of missing values under each variable
missing_values <- population_data %>%
  summarise_all(~sum(is.na(.)))  %>%
  gather(variable, missing_count) %>%
  filter(missing_count > 0)
missing_values

## As the rows are all zero, we have no missing values
```

```{r}
# Cleaning variable names for easier handling
names(population_data) <- make.names(names(population_data))
head(population_data, 2) #checking whether the variable names changed
```

## 2.3 Checking for Outliers 

```{r}
View(population_data)
```


```{r outliers}
# Selecting quantitative variables to identify outliers
numeric_vars <- population_data %>%
  select(X2022.Population, Area..km..., Density..per.km..., Growth.Rate, World.Population.Percentage) %>%
  gather(variable, value) %>%
  filter(!is.na(value))
numeric_vars
```


```{r}
# Visualizing outliers using box plot
ggplot(numeric_vars, aes(x = variable, y = value)) +
  geom_boxplot(fill = "lightblue", alpha = 0.7) +
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  labs(title = "Box Plots for Outlier Detection",
       x = "Variables",
       y = "Values") +
  theme(axis.text.x = element_blank())
```
We shall not deal with outliers because in instances such as population, it is showing data diversity in different countries rather than errors in data.


# 3. Population Projection for 2030

Using the formula: Nt = P * e^(r*t)

```{r population_projection}
# Calculating population projection for 2030
# P = 2022 population, r = growth rate, t = 8 years (2030-2022)
population_data <- population_data %>%
  mutate(
    Population_2030 = X2022.Population * exp(Growth.Rate * 8),
    Population_2030 = round(Population_2030)
  )
```


```{r}
# Displaying top 10 populous countries with their projections
top10_2022 <- population_data %>%
  arrange(desc(X2022.Population)) %>%
  slice_head(n = 10) %>%
  transmute(
    Rank = row_number(),
    `Country / Territory` = Country.Territory,
    `Population (2022)`   = X2022.Population,
    `Projected 2030`      = Population_2030
  )

tbl_gt <- top10_2022 %>%
  gt() %>%
  tab_caption("Top 10 Most Populous Countries (2022) with 2030 Projections") %>%
  fmt_number(
    columns = c(`Population (2022)`, `Projected 2030`),
    decimals = 0, use_seps = TRUE
  )
tbl_gt
gt::gtsave(tbl_gt, "top10_2022_with_proj.png",
           vwidth = 1100, vheight = 500, expand = 5)
```


# 4. Value Extraction and Visualization

## 4.1 Top 10 Most Populous Countries in 2022

```{r top10_countries}
# Extracting top 10 most populous countries
top10_pop <- population_data %>%
  arrange(desc(X2022.Population)) %>%
  head(10)
top10_pop
```

```{r}
# Using ggplot2 to plot bar graph of top 10 most populous countries
ggplot(top10_pop, aes(x = reorder(Country.Territory, X2022.Population), y = X2022.Population/1e6)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  coord_flip() +
  labs(title = "Top 10 Most Populous Countries (2022)",
       x = "Country",
       y = "Population (Millions)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))
```


```{r}
# displaying the top 10 populous countries in a table
top10_pop %>%
  select(Country.Territory, X2022.Population) %>%
  mutate(Population_Millions = round(X2022.Population/1e6, 2)) %>%
  kable(caption = "Top 10 Most Populous Countries (2022)")
```

## 4.2 Population Trend (1990-2022) for Top 10 Countries

```{r population_trend}
# Preparing data for trend analysis
top10_countries <- top10_pop$Country.Territory

trend_data <- population_data %>%
  filter(Country.Territory %in% top10_countries) %>%
  select(Country.Territory, X1990.Population, X2000.Population, X2010.Population, 
         X2015.Population, X2020.Population, X2022.Population) %>%
  gather(Year, Population, -Country.Territory) %>%
  mutate(Year = as.numeric(gsub("X([0-9]+)\\.Population", "\\1", Year)))
trend_data
```

```{r}
# Line plot showing population trends
ggplot(trend_data, aes(x = Year, y = Population/1e6, color = Country.Territory)) +
  geom_line(size = 1.2, alpha = 0.8) +
  geom_point(size = 2) +
  labs(title = "Population Trend (1990-2022) for Top 10 Most Populous Countries",
       x = "Year",
       y = "Population (Millions)",
       color = "Country") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_color_viridis_d()
```


## 4.3 CO2 Emissions Trend for Top 10 Countries

```{r}
#viewing CO2 emissions dataset
co2_data
```

```{r co2_trend}
# Extracting clean data set of top 10 countries from CO2 data set also found in top10_countries data set
co2_clean <- co2_data %>%
  filter(Country.Name %in% top10_countries) %>%
  select(Country.Name, X1990:X2019.1) %>%
  gather(Year, CO2_Emission, -Country.Name) %>%
  mutate(
    Year = as.numeric(gsub("X([0-9]+).*", "\\1", Year)),
    CO2_Emission = as.numeric(CO2_Emission)
  ) %>%
  filter(!is.na(CO2_Emission), Year >= 1990, Year <= 2019)
co2_clean
```


```{r}
# Visualizing Line plot for CO2 emissions trend
ggplot(co2_clean, aes(x = Year, y = CO2_Emission, color = Country.Name)) +
  geom_line(size = 1.2, alpha = 0.8) +
  geom_point(size = 1.5) +
  labs(title = "CO2 Emissions Trend (1990-2019) for Top 10 Most Populous Countries",
       x = "Year",
       y = "CO2 Emissions (metric tons per capita)",
       color = "Country") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_color_viridis_d()
```

# 5. Correlation Analysis

## 5.1 Correlation Matrix for Population Variables

```{r correlation_analysis}
# Selecting variables for correlation analysis
corr_vars <- population_data %>%
  select(Area..km..., Density..per.km..., Growth.Rate, World.Population.Percentage) %>%
  na.omit()
corr_vars
```

```{r}
# Calculating correlation matrix using  numerical values
corr_matrix <- cor(corr_vars)
cat("Correlation Matrix (Numerical Values):\n")
print(round(corr_matrix, 3))
```

```{r}
# Correlation heatmap
corrplot(corr_matrix, method = "color", type = "upper", 
         order = "hclust", tl.cex = 0.8, tl.col = "black",
         title = "Correlation Heatmap for Population Variables",
         mar = c(0,0,2,0))
```


## 5.2 Population vs CO2 Emissions Analysis

```{r pop_co2_correlation}
# Merge datasets for correlation analysis
# First, clean country names for matching
population_2022 <- population_data %>%
  select(Country.Territory, X2022.Population) %>%
  rename(Country = Country.Territory, Population_2022 = X2022.Population)

co2_2019 <- co2_data %>%
  select(Country.Name, X2019.1) %>%
  rename(Country = Country.Name, CO2_2019 = X2019.1) %>%
  mutate(CO2_2019 = as.numeric(CO2_2019)) %>%
  filter(!is.na(CO2_2019))

# Merging datasets
merged_data <- inner_join(population_2022, co2_2019, by = "Country")
merged_data
```


```{r}
# Scatter plot
ggplot(merged_data, aes(x = Population_2022/1e6, y = CO2_2019)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Relationship between Population (2022) and CO2 Emissions (2019)",
       x = "Population 2022 (Millions)",
       y = "CO2 Emissions 2019 (metric tons per capita)") +
  theme_minimal()
```

```{r}
# Calculate correlation
correlation_coef <- cor(merged_data$Population_2022, merged_data$CO2_2019, use = "complete.obs")

cat("Correlation coefficient between Population and CO2 Emissions:", round(correlation_coef, 3))
```

```{r}
# Linear regression analysis
lm_model <- lm(CO2_2019 ~ Population_2022, data = merged_data)
cat("\n\nLinear Regression Summary:\n")
summary(lm_model)
```

# 6. Continental CO2 Emissions Comparison

```{r continental_co2}
# First, we need to add continent information to CO2 data
# Get continent mapping from population data
continent_mapping <- population_data %>%
  select(Country.Territory, Continent) %>%
  rename(Country = Country.Territory)

# Merge with CO2 data
co2_with_continent <- co2_data %>%
  left_join(continent_mapping, by = c("Country.Name" = "Country")) %>%
  filter(!is.na(Continent))
co2_with_continent$Continent
```

```{r}
# Calculating the total CO2 emissions by continent for 2019
continental_co2 <- co2_with_continent %>%
  mutate(CO2_2019 = as.numeric(X2019.1)) %>%
  filter(!is.na(CO2_2019)) %>%
  group_by(Continent) %>%
  summarise(
    Total_CO2_2019 = sum(CO2_2019, na.rm = TRUE),
    Countries_Count = n(),
    .groups = 'drop'
  ) %>%
  arrange(desc(Total_CO2_2019))


continental_co2 %>% kable(caption = "CO2 Emissions by Continent (2019)")

```


```{r}
# Bar plot of continental CO2 emissions
ggplot(continental_co2, aes(x = reorder(Continent, Total_CO2_2019), y = Total_CO2_2019)) +
  geom_col(fill = "coral", alpha = 0.8) +
  coord_flip() +
  labs(title = "Total CO2 Emissions by Continent (2019)",
       x = "Continent",
       y = "Total CO2 Emissions (metric tons per capita)") +
  theme_minimal()
```


```{r}
# Answer the specific questions
cat("1. First continent with highest 2019 CO2 emission:", continental_co2$Continent[1])
cat("\n2. Third continent in terms of emitting CO2:", continental_co2$Continent[3], 
    "with total emission:", round(continental_co2$Total_CO2_2019[3], 2), "metric tons per capita")
cat("\n3. Last continent in terms of emitting CO2:", tail(continental_co2$Continent, 1), 
    "with total emission:", round(tail(continental_co2$Total_CO2_2019, 1), 2), "metric tons per capita")
```

# 7. Summary Statistics

```{r summary_stats}
# Summary statistics for key variables
cat("Summary Statistics for Population Dataset:\n")
population_data %>%
  select(X2022.Population, Area..km..., Density..per.km..., Growth.Rate) %>%
  summary()
```

```{r}
cat("\n\nSummary Statistics for Merged Dataset:\n")
merged_data %>%
  select(Population_2022, CO2_2019) %>%
  summary()
```


# 8. Key Findings and Insights

## Key Findings:

1. **Population Analysis**: The analysis reveals significant variations in population sizes, with China and India leading as the most populous countries.

2. **Population Projection**: Using the exponential growth model, countries with higher growth rates show substantial projected increases by 2030.

3. **CO2 Emissions Patterns**: There's a complex relationship between population size and per capita CO2 emissions, suggesting that factors beyond population influence emission levels.

4. **Continental Differences**: Different continents show varying patterns of CO2 emissions, reflecting economic development levels and energy consumption patterns.

5. **Correlation Insights**: The correlation analysis between population variables reveals interesting relationships between area, density, growth rate, and world population percentage.

## Data Quality Notes:

- The datasets were found to be relatively clean with minimal missing values
- Some outliers were identified in the quantitative variables, particularly in area and density measurements
- Country name matching between datasets required careful handling due to slight variations in naming conventions

## Recommendations:

1. Further analysis could include economic indicators to better understand CO2 emission patterns
2. Time-series analysis could provide more insights into emission trends
3. Regional analysis within continents could reveal more granular patterns

---

*END*